Management induced changes of soil organic carbon on global croplands

### 2022-02-28 SOCbudgetPaper2/discussion.Rmd
Major update of finalized Paper - Draft 2

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% copernicus_rticles_template (flag for rticles template detection - do not remove!)

%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.

%% 2-column papers and discussion papers
\documentclass[gc, manuscript]{copernicus}

%% Journal abbreviations (please use the same for preprints and final revised papers)

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% EGUsphere (egusphere) | This is only for EGUsphere preprints submitted without relation to an EGU journal.
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Bone and Joint Infection (jbji)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Polarforschung - Journal of the German Society for Polar Research (polf)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Safety of Nuclear Waste Disposal (sand)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

% Pandoc citation processing

% The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
% % % From pandoc table feature
% \usepackage{longtable,booktabs,array}
% % \usepackage{calc} % for calculating minipage widths
% % Correct order of tables after \paragraph or \subparagraph
% \usepackage{etoolbox}
% \makeatletter
% \patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
% \makeatother
% % Allow footnotes in longtable head/foot
% \IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
% \makesavenoteenv{longtable}
% 
% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%
\begin{document}

\title{Management induced changes of soil organic carbon on global croplands}

\Author[1, 3]{Kristine}{Karstens}
\Author[1]{Benjamin Leon}{Bodirsky}
\Author[1]{Jan Philipp}{Dietrich}
\Author[2]{Marta}{Dondini}
\Author[1]{Jens}{Heinke}
\Author[2]{Matthias}{Kuhnert}
\Author[1]{Christoph}{Müller}
\Author[1]{Susanne}{Rolinski}
\Author[2]{Pete}{Smith}
\Author[1]{Isabelle}{Weindl}
\Author[1, 3]{Hermann}{Lotze-Campen}
\Author[1]{Alexander}{Popp}

\affil[1]{Potsdam Institute for Climate Impact Research (PIK), Member of the Leibniz Association, P.O. Box 60 12 03, 14412 Potsdam, Germany}
\affil[2]{Institute of Biological \& Environmental Sciences, University of Aberdeen, Aberdeen, UK}
\affil[3]{Humboldt-Universität zu Berlin, Department of Agricultural Economics, Unter den Linden 6, 10099 Berlin, Germany}

\runningtitle{R Markdown Template for Copernicus}

\runningauthor{Nüst et al.}

\correspondence{Kristine\ Karstens\ (\href{mailto:kristine.karstens@pik-potsdam.de}{\nolinkurl{kristine.karstens@pik-potsdam.de}})}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. The first meter of the Earth's soils profile stores three times as much carbon as the vegetation and twice the amount of C in the atmosphere. SOC has been depleted by anthropogenic land-cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatially explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with a reduced-complexity model based on the IPCC Tier 2 method to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around 34.6 GtC relative to a counterfactual hypothetical natural state in 1975. Within the period 1975--2010 this SOC debt continued to expand by 5 GtC (\(0.14\unit{GtC yr^{-1}}\)) to around 39.6 GtC. However, accounting for historical management led to 2.1 GtC less (\(0.06\unit{GtC yr^{-1}}\)) emissions than under the assumption of constant management. We also find that management decisions have influenced the historical SOC trajectory most strongly by residue returning, indicating that increasing SOC sequestration by biomass retention may be a promising negative emissions technique. The reduced-complexity SOC model may allow to simulate management-induced SOC sequestration also within computationally demanding integrated (land-use) assessment modeling.
\end{abstract}

\newpage

\introduction

Soil organic carbon (SOC), the amount of organic carbon stored in the Earth's soil, constitutes the largest terrestrial organic carbon pool. It exceeds the carbon in the atmospheric and vegetation pools multiple times \citep{batjes_total_1996}. Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO\textsubscript{2} in the atmosphere \citep{frielingstein_glob_carbon_2020, minasny_soil_2017}. The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile \citep{batjes_total_1996, sanderman_soil_2017}.

Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time. Recent studies have focused on the evaluation of total SOC stocks of the world as well as on the spatial disaggregation of soil properties such as SOC content \citep{batjes_harmonized_2016, hengl_soilgrids250m_2017, fao_global_2018}. However, these studies often do not include human interventions, like land cover change and agricultural management, in their analysis. Compared to climatic and geological driving forces, human interventions alter terrestrial carbon pools over much shorter time scales and are currently one of the most dominant drivers of SOC changes on managed land \citep{hansis_relevance_2015, bastos_comparison_2021}.

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions, see Pongratz et al.~\citeyearpar{pongratz_terminolgy_2014}), which is the amount of organic carbon soils have lost under cultivation compared to a hypothetical potential natural vegetation state. Sanderman et al.~\citeyearpar{sanderman_soil_2017} identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter \citep{lal_world_2001}.

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs), Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions \citep{frielingstein_glob_carbon_2020}. While providing estimates of the magnitude of SOC losses due to land-cover change, most models lack a detailed representation of agricultural management. Earlier DGVM- and ESM-based assessments only considered changes in land cover, but ignored the removal of biomass at harvest \citep{strassmann_simulating_2008, betts_climate_2015}. BKMs are designed to estimate LUC related emissions but often ignore changes in SOC due to climate change, CO\textsubscript{2} fertilization and N deposition. Whereas BKM have largely improved in estimating additional emissions from wood harvest and shifting cultivation, state of the art models do not consider impacts of varying agricultural management \citep{frielingstein_glob_carbon_2020, houghton_carbon_2012, hansis_relevance_2015, bastos_comparison_2021}.

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance \citep[e.g.~][]{bondeau_modelling_2007, lindeskog_implications_2013}. Pugh et al.~\citeyearpar{pugh_simulated_2015} explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with stylized scenarios rather than with historical management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historical LUC and the slow decline of SOC under cropland before a new equilibrium is reached.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time (e.g.~growing seasons \citep{portmann_mirca2000global_2010}, multiple cropping systems \citep{waha_multiple_2020}, irrigation systems \citep{jagermeyr_water_2015}) or as stylized scenarios \citep[e.g.~][]{pugh_simulated_2015, lutz_simulating_2019}. Herzfeld et al.~\citeyearpar{herzfeld_soc_2021} account for historical changes in fertilizer and manure inputs, residue removal rates and tillage systems and report SOC losses from cropland expansion over the period from 1700--2018 of 215 GtC. Within their stylized future management scenarios they find that none of the management aspects considered (residue management, no-tillage) can create a net carbon sink on current cropland areas under future climate change.

More data sets on spatially explicit agricultural management time series with global coverage are becoming available (e.g.~on tillage systems, see \citep{porwollik_generating_2019, prestele_spatially_2018}) and modeling approaches are increasingly being developed to project the dynamics of management systems into the future (e.g.~\citep{iizumi_modeling_2019, minoli_modelling_2019}), but have --- to our knowledge --- not yet found their way into comprehensive assessments of the terrestrial carbon cycle in DGVMs and BKMs.

Field-scale models \citep{del_grosso_simulated_2001, coleman_simulating_1997, smith_estimating_2010, taghizadeh-toosi_c-tool_2014} are able to better account for historical agricultural management if detailed information on crop yield levels, fertilizer inputs and various other on-farm measures is available for the studied sites. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global domain remains a complex challenge \citep{morais_detailed_2019}.

Managed soils have been increasingly studied not only for their carbon emitting behavior, but also because of their capacity to re-store carbon (soil carbon sequestration (SCS) techniques). However, assessing SCS dynamically considering the interdependency with environmental, social and economic sustainability targets has been difficult so far, as integrated assessment models (IAMs) \citep{popp_land-use_2016, rogelj_scenarios_2018} have not integrated soil management into their mitigation pathways. More detailed process-based models are typically computationally too demanding to be integrated into optimization-based IAMs. Better accounting for soil carbon management in IAMs thus requires a light-weight model suitable for iterative modeling with detailed options to represent agricultural soil management.

The objectives of our study are (1) to develop a reduced-complexity SOC model able to account for SCS in IAM frameworks; (2) to create a comprehensive data set of the global gridded management time series, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices; and (3) to provide global as well as spatially explicit SOC and SOC debt estimates that consider spatially explicit and time-variant agricultural management.
We decompose the contribution of different management activities through a scenario analysis, identifying the most impacting management decisions for SOC development. Moreover, we compare our model performance against other SOC stock and SOC emission estimates, to evaluate the suitability of this reduced-complexity approach for integration into IAM modeling.
\newpage

\hypertarget{methods}{%
\section{Methods}\label{methods}}

In Sect. \ref{sec:carbonbudget} we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail in the refinement of the IPCC guidelines vol.~4 Chapter 5 on ``Cropland'' \citep{ogle_cropland_in_ipcc_2019}. We additionally describe how we configured and extended the Tier 2 modeling approach \citep[for model code see][]{karstens_mrsoil_2020}. In Sect. \ref{sec:tier1} we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines \citep{eggleston_ipcc_2006, calvo_buendia_ipcc_2019}.
Section \ref{sec:agrimanagement} provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices \citep[for model code see][]{bodirsky_mrcommons_2020}. In Sect. \ref{sec:scenarios} we define the management scenarios used to analyze the role of different management aspects in historical cropland SOC dynamics.

\hypertarget{sec:carbonbudget}{%
\subsection{SOC stocks and stock changes following the Tier 2 modeling approach}\label{sec:carbonbudget}}

Following the Tier 2 modeling approach of the refinement of the IPCC guidelines vol.~4 Chapter 5 on ``Cropland'' \citep{ogle_cropland_in_ipcc_2019}; referred to as \textit{Tier 2 modeling approach} in the following), we estimate SOC stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current SOC state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period:
(1) We calculate annual land-use type-specific steady states and decay rates for \(SOC\) stocks (Sect. \ref{sec:steadystates});
(2) we account for land conversion by transferring \(SOC\) from and to natural vegetation (Sect. \ref{sec:carbontransfer}),
(3) we estimate \(SOC\) stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \ref{sec:totalsoc}).
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \ref{sec:initsoc}).

\hypertarget{sec:steadystates}{%
\subsubsection{Steady-state SOC stocks and decay rates}\label{sec:steadystates}}

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the Tier 2 modeling approach for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model \citep{parton_analysis_1987}. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below- and above-ground crop residues left or returned to the field (see Sect. \ref{sec:residues}) and manure inputs (see Sect. \ref{sec:livstmanure}); for natural vegetation, litterfall including fine root turnover \citep{schaphoff_lpjml4_2018} is the only source of carbon inflow to the soil. Following the IPCC guidelines \citep{ogle_cropland_in_ipcc_2019}, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective \(SOC\) sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools.

Whereas residue and manure default lignin and nitrogen fractions are given by the IPCC guidelines \citep{ogle_cropland_in_ipcc_2019}, we use plant-functional type and plant-organ specific parameterization for the natural litterfall. Global distribution of plant-functional types is given by \citep{schaphoff_lpjml4_2018} as well as separation of litter into leaf, fine root and wood litter compartments excluding litter biomass burnt in wild fires. Leaf litter parameters are given by Brovkin et al.~\citeyearpar{brovkin_plant-driven_2012}, fine root to leaf litter lignin ratio by Guo et al.~\citeyearpar{guo_leaf_root_decompostion_2021}, lignin content of wood litter by Rahman et al.~\citeyearpar{rahman_lignin_2013} and nitrogen content scaling factors for leaf to fine roots and leaf to wood litter by von Bloh et al.~\citeyearpar{vBloh_lpjml5_2018}.
Data sources for all considered carbon inputs as well as for lignin and nitrogen parameterization are listed in Table \ref{tab:datasourceinputs}.

  \begin{table*}[h]
  \caption{type and data sources for carbon inputs and parameterization to different land-use types }
  \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.20\textwidth} p{0.35\textwidth}}
  \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
  \middlehline
  \multirow{3}{*}{cropland} & above-ground residues, & \multirow{3}{*}{\begin{minipage}[t]{0.35\columnwidth}\raggedright
                                                                      \citet{faostat_faostat_2016}, \\
                                                                      \citet{schaphoff_lpjml4_2018},\\
                                                                      \citet{weindl_livestock_2017} \end{minipage}} & 
                         \multirow{3}{*}{\begin{minipage}[t]{0.35\columnwidth}\raggedright
                         LG:C generic values according to Table 5.5B,\\
                         5.5C from IPCC \citep{ogle_cropland_in_ipcc_2019},\\
                         crop-specific N:C from \citet{bodirsky_n2o_2012}
                                                                      \end{minipage}} \\
                            & below-ground residues, &  &  \\
                            & manure                &  &  \\ 
                            \hline
  natural vegetation        & annual litterfall &  \citet{schaphoff_lpjml4_2018}
                      & \begin{minipage}[t]{0.35\columnwidth}\raggedright\strut 
                          leaf N and LG concentration from \citet{brovkin_plant-driven_2012}, 
                          root to leaf litter LG ratio \citet{guo_leaf_root_decompostion_2021},  
                          lignin content of wood litter \citet{rahman_lignin_2013} and 
                          nitrogen scaling factors for leaf to root and wood litter 
                          from \citet{vBloh_lpjml5_2018} \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

The sub-pool specific decay rates \(k_{sub}\) are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over space \(i\) and time \(t\). Following the Tier 2 modeling approach \citep{ogle_cropland_in_ipcc_2019}, we consider temperature \(temp\), water \(wat\), sand-fraction \(sf\), and tillage \(till\) effects to account for spatial and temporal variation of decay rates. Thus, \(k_{sub}\) rates are given by

\begin{equation}
\begin{aligned}
& k_{i,t,\mathrm{active},lu}  & = &~ k_{\mathrm{active}}  ~ &\cdot~ temp_{i,t} ~ &\cdot~ wat_{i,t,lu} ~ &\cdot~ till_{i,t,lu} ~ & \cdot~ sf_{i}\\
& k_{i,t,\mathrm{slow},lu}    & = &~ k_{\mathrm{slow}}    ~ &\cdot~ temp_{i,t} ~ &\cdot~ wat_{i,t,lu} ~ &\cdot~ till_{i,t,lu} ~ &\\
& k_{i,t,\mathrm{passive},lu} & = &~ k_{\mathrm{passive}} ~ &\cdot~ temp_{i,t} ~ &\cdot~ wat_{i,t,lu} ~ & ~ & 
\label{eq:decayrates}
\end{aligned}.
\end{equation}

For natural vegetation, we assume rainfed and non-tilled conditions, whereas for cropland, we distinguish the effect of different tillage (see Sect. \ref{sec:tillage}) and irrigation (see Sect. \ref{sec:irrigation}) practices on decay rates. We calculate area-weighted means for \(till\) and \(wat\) on cropland for each grid cell, using area shares for the different tillage and irrigation practices. Data sources as well as used parameters for the different decay drivers for all land-use types are listed in Table \ref{tab:datasourcedecay}; equations are displayed by equations 5.0B--5.0F in Ogle et al.~\citeyearpar{ogle_cropland_in_ipcc_2019}.

 \begin{table*}[h]
 \caption{type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & soil quality & sand fraction of the first 0-30 cm 
                                     & \citet{hengl_soilgrids250m_2017} \\
                      \cline{2-4}

                      & mircobial activity & air temperature & \citet{harris_version_2020} \\
                      \cline{2-4}

                      & soil moisture & precipitation \& potential evapotranspiration & \citet{harris_version_2020} \\
                      \cline{1-4}
\multirow{2}{*}{\begin{minipage}[t]{0.15\columnwidth}\raggedright\strut cropland\\(additionally)\strut\end{minipage}} & soil moisture*  & irrigation  & Sect. \ref{sec:irrigation} \\ 
                      \cline{2-4}

                      & soil disturbance & tillage & Sect. \ref{sec:tillage} \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 \label{tab:datasourcedecay}
 \end{table*}

\hypertarget{sec:carbontransfer}{%
\subsubsection{SOC transfer between land-use types}\label{sec:carbontransfer}}

\begin{figure}[h]
\includegraphics[width=8cm]{../ResultNotebooks/Output/Images/LUtransitions} \caption{Scheme of land-use transition representation. 
Given an initial land-use pattern (as in this example 2 ha land under natural vegetation and 1 ha of cropland), there are separate $SOC$ stocks for natural vegetation and cropland. While in this example we assume $SOC$ under natural vegetation to be in steady state, the cropland $SOC$ stock approaches its steady state without having reached it yet (a). Upon cropland expansion (in this example half of the natural vegetation is cleared to be used as cropland), $SOC$ stocks on cropland increase due to a transfer of land from natural vegetation (b). Separately accounting for SOC dynamics on newly converted and existing cropland (c) leads to the same overall SOC dynamics as if SOC stocks are averaged (d), due to the linearity of Eq. \ref{eq:SOCstate} and the cropland-age independend decay rates (see Eq. \ref{eq:decayrates})}\label{fig:LUtrans}
\end{figure}

We calculate SOC stocks based on the area shares of land-use types \(lu\) within our grid cells \(i\). If land is converted from one land-use type \(lu=\{crop,natveg\}\) into the other \(!lu=\{natveg,crop\}\), a respective share of the SOC is reallocated within our budget. We do not distinguish between newly converted and existing cropland, but can work with the average carbon content as the relative decay of SOC is proportional to the SOC stock (see \ref{fig:LUtrans}). We account for land conversion at the beginning of each time step \(t\) by calculating a preliminary stock \(SOC_{t^*}\) via

\begin{equation}
SOC_{i,t^*,sub,lu} = SOC_{i,t-1,sub,lu} - \frac{SOC_{i,t-1,sub,lu}}{A_{i,t-1,lu}} \cdot  AR_{i,t,lu} + \frac{SOC_{i,t-1,sub,!lu}}{A_{i,t-1,!lu}} \cdot  AE_{i,t,lu}
\label{eq:ctransfer}
\end{equation}

with \(A_{lu}\) being the land-use type specific areas, \(AR_{lu}\) the area reduction and \(AE_{lu}\) the area expansion of the two land-use types. Data sources and methodology on land-use states and changes are described in Sect. \ref{sec:landuse}.

\hypertarget{sec:totalsoc}{%
\subsubsection{Total SOC stocks and stock changes}\label{sec:totalsoc}}

SOC converges towards the calculated steady-state stock \(SOC^{\mathrm{eq}}\) for each grid cell \(i\), each annual time step \(t\), each land-use type \(lu\) and each sub-pool \(sub\) like

\begin{equation}
SOC_{i,t,sub,lu} = SOC_{i,t^*,sub,lu} + (SOC^{\mathrm{eq}}_{i,t,sub,lu} - SOC_{i,t^*,sub,lu}) \cdot k_{i,t,sub,lu} \cdot 1\unit{a}.
\label{eq:SOCstate}
\end{equation}

Note that the decay rates have to be multiplied by one year (\(1a\)) to form a dimensionless factor.
Reformulating this equation, we obtain a mass balance equation as follows

\begin{equation}
SOC_{i,t,sub,lu} = SOC_{i,t^*,sub,lu} - \underbrace{SOC_{i,t^*,sub,lu} \cdot k_{i,t,sub,lu} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{\mathrm{eq}}_{i,t,sub,lu} \cdot k_{i,t,sub,lu} \cdot 1\unit{a}}^{\text{input (using equation (1))}}.
\label{eq:steadystate2budget}
\end{equation}

The global \(SOC\) stock for each time step \(t\) can then be calculated via

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lu} \overbrace{\sum_{sub} SOC_{i,t,sub,lu}}^{\text{$SOC_{i,t,lu}$ -- land-use type specific $SOC$ stock within cell}}}_{\text{$SOC_{i,t}$ -- total $SOC$ stock within cell}}.
\label{eq:totalstock}
\end{equation}

According to the IPCC guidelines \(SOC\) changes can be expressed as the difference of two consecutive years \citep[see Eq. 5.0A in][]{ogle_cropland_in_ipcc_2019}. This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we define the absolute and relative \(SOC\) changes in relation to a potential natural state \(SOC^{\mathrm{pnv}}\) under the same climatic conditions in grid cell \(i\) at time \(t\), that is based on the natural vegetation \(SOC\) calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes \(\Delta SOC\) and relative changes \(F^{\mathrm{SCF}}\) are thus given by

\begin{equation}
\Delta SOC_{i,t} = SOC_{i,t} - SOC^{\mathrm{pnv}}_{i,t}\qquad \text{and} \qquad  F^{\mathrm{SCF}}_{i,t} = \frac{SOC_{i,t}}{SOC^{\mathrm{pnv}}_{i,t}} .
\label{eq:stockdiff}
\end{equation}

Note that the absolute changes \(\Delta SOC\) can be also interpreted as the SOC debt \citep{sanderman_soil_2017} due to human cropping activities; whereas relative changes \(F^{\mathrm{SCF}}\) can be considered stock change factors as defined within the IPCC guidelines of 2006 \citep{eggleston_ipcc_2006}. Moreover, \(\Delta SOC\) is equivalent to the negated cumulative SOC component of human land-use change emissions \citep{pugh_simulated_2015}.

\hypertarget{sec:initsoc}{%
\subsubsection{Initialization of SOC pools}\label{sec:initsoc}}

The initialization of \(SOC\) pools is very important and has to include the proper accounting for the land-use history, as many CO\textsubscript{2} emissions from agricultural soils are caused by historical land-use change (LUC) and the slow decline of \(SOC\) under crop cultivation, before it reaches a new equilibrium.
We initialize our \(SOC\) sub-pools using a three-step approach, since input data availability is limited for climate and litter estimates (starting only in 1901) as well as for agricultural management data (starting only in 1965):

Firstly, in order to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we consider land-use change from 1510 onwards. In 1510, we assume all \(SOC\) pools to be in natural steady-state, implying that all land-use change prior to that time occurs in 1510. We assume that by 1901, all cropland converted in 1510 has reached its new steady state, so that it is not necessary to explicitly account for even older land conversion. Model inputs for 1901--1930 for climate and natural vegetation litterfall are repeated for 1510--1900 to mimic constant climate conditions for this first initializing period. Similarly, agricultural management data are held constant at the level of 1965 until 1965. This approach follows others studies looking on effects of land-use change and management \citep[e.g.~][]{schaphoff_lpjml4_2018-1, herzfeld_soc_2021}.

Secondly, with the availability of transient climate data after 1901, we account not only for land-use change, but also for historical climate change and consequently natural litter inputs to the soil from 1901 to 1965 still considering constant agricultural input data, which are not available prior to 1965.

Thirdly, we run the model for 10 years from 1965 to 1975 with historical dynamic data on agricultural management and start analyzing results from 1975 onward. This is in line with the IPCC guidelines vol.~4 method suggestion to have a 5-20 year spin-up period \citep{ogle_cropland_in_ipcc_2019}.

With transient climate considered after 1901, decay rates \(k_{sub}\) become dynamic in time. As the decay rates are also affected by irrigation and tillage (see Sect. \ref{sec:steadystates}), we also account for transient changes in irrigated areas after 1901. Data on no-tillage practices are only available after 1974 and we assume conventional tillage on all cropland prior to 1975.

\hypertarget{sec:tier1}{%
\subsection{SOC stocks and stock changes following Tier 1}\label{sec:tier1}}

Additionally to the Tier 2 modeling approach \citep{ogle_cropland_in_ipcc_2019} and the detailed analysis of management data coming with it, \(SOC\) changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines \citep{eggleston_ipcc_2006, calvo_buendia_ipcc_2019}. Here, stocks are calculated via stock change factors (\(F^{\mathrm{SCF}}\)) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of \(F^{\mathrm{SCF}}\) are differentiated by crops, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these flows. Moreover, estimates of \(F^{\mathrm{SCF}}\) vary for different climatic zones (\(c\)) specified by the IPCC (see Fig. \ref{fig:CLIMzone}). The actual \(SOC\) stocks are thus calculated based on a given reference stock \(SOC^{\mathrm{ref}}\) by

\begin{equation}
SOC_{i,t} = \sum_{c,m} T_{c,i} \cdot SOC^{\mathrm{ref}}_{i,t} \cdot F^{\mathrm{SCF}}_{c,m}
\label{eq:tier1}
\end{equation}

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into corresponding climate zones \(c\). For this analysis, we use the default \(F^{\mathrm{SCF}}\) from the Tier 1 method of Eggleston et al.~\citeyearpar{eggleston_ipcc_2006} and Calvo Buendia et al.~\citeyearpar{calvo_buendia_ipcc_2019} as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

\hypertarget{sec:agrimanagement}{%
\subsection{Agricultural management data at 0.5 degree resolution}\label{sec:agrimanagement}}

We compile country-specific FAO production and cropland statistics \citep{faostat_faostat_2016} to a harmonized and consistent data set. The data is prepared in 5-year time steps from 1965 to 2010, which restricts our analysis to the time span from 1975 to 2010 (after a spin-up phase from 1510--1974). For all the following data, if not declared differently, we interpolate values linearly between the time steps and keep them constant before 1965.

\hypertarget{sec:landuse}{%
\subsubsection{Land use and land-use change}\label{sec:landuse}}

Land-use patterns are based on the Land-Use Harmonization 2 \citep{hurtt_harmonization_2020} data set (short LUH2), which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area (in Mha) of the five different cropland subcategories (c3ann: C3 annual crops, c3per: C3 perennial crops, c4ann: C4 annual crops, c4per: C4 perennial crops, c3nfx: C3 nitrogen-fixing crops) of LUH2 into our 17 crop groups (see Table \ref{tab:fao2luh2mag}), applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data \citep{faostat_faostat_2016}. By calculating country-specific multicropping factors \(MCF\) using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level.
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

\hypertarget{sec:residues}{%
\subsubsection{Crop and crop residues production}\label{sec:residues}}

Crop production patterns are compiled crop group specific using half-degree yield data from LPJmL \citep{schaphoff_lpjml4_2018} as well as half-degree cropland patterns (see Sect. \ref{sec:landuse}). We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production \citep{faostat_faostat_2016}. Note

Crop residue production and management is based on a revised methodology of Bodirsky et al.~\citeyearpar{bodirsky_n2o_2012} and key aspects are explained as they play a central role in soil carbon modeling. Starting from crop yield estimates \(Y\) and respective physical crop area \(CA\), we estimate total above-ground \(AGR\) and below-ground \(BGR\) residue biomass (in tonnes) using crop group \(cg\) specific ratios for above-ground residues to harvested biomass \(r^{\mathrm{ag,prod}}_{cg}\) in \(\unit{tC\,\unit{tC}^{-1}}\), above-ground residues to harvested area \(r^{\mathrm{ag,area}}_{cg}\) in \(\unit{tC\,\unit{ha}^{-1}}\) and below-ground residues to above-ground biomass \(r^{\mathrm{bg}}_{cg}\) in \(\unit{tC\,\unit{tC}^{-1}}\) as follows

\begin{equation}
\begin{aligned}
AGR_{i,t,cg} & = CA_{i,t,cg} \cdot \left( Y_{i,t,cg} \cdot  r^{\mathrm{ag,prod}}_{cg} + MCF_{i,t} \cdot r^{\mathrm{ag,area}}_{cg} \right)
\qquad & \textrm{and} \\
BGR_{i,t,cg} & = (CA_{i,t,cg} \cdot Y_{i,t,cg} + AGR_{i,t,cg}) \cdot r^{\mathrm{bg}}_{cg} \quad.
\label{eq:resbiomass}
\end{aligned}
\end{equation}

>>>>

# Discussion
